---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading and preprocessing the data
For loading the data, you need to set your working directory in the same location as the file and then use the command "read.csv" as follows:

```{r}
tabla <- read.csv("./activity.csv")
```

As you can see, I assigned the table information to the variable "tabla". We can see the content of tabla using "head":

```{r}
head(tabla)
```

As you can see, the variable contains the require information. To continue working with the table, we will erase the NA values in a new variable called "tablanoNAs" and we will check the content using the same head command.

```{r}
tablanoNAs <- na.omit(tabla)
head(tablanoNAs)
```

You can confirm the NA deletion using the next command:

```{r}
sum(is.na(tabla$steps))

sum(is.na(tablanoNAs$steps))
```

You will notice that there are no more NA rows in the steps column. So now we can process our data.

## What is mean total number of steps taken per day?

For this, we need to assign the date column as a factor to calculate the mean by day. We can use the dplyr Package to group the data and assign it to a new variable. In the new table, the column with the sum of steps for each day will be called "Total_steps".

```{r}
library(dplyr, warn.conflicts = FALSE)

tablaporfecha <- tablanoNAs %>% group_by(date) %>% summarise(Total_steps = sum(steps))

head(tablaporfecha)
```

Once the table is grouped, we can generate an histogram using the base ploting system. Lets modify the breaks in the histogram so we can have a better idea of how is the information distributed across the dates:

```{r}
hist(tablaporfecha$Total_steps, xlab = "Total steps", main = "Histogram of the total number of steps taken each day", breaks = 10)
```

Furthermore, the mean and median of the steps taken by day can be found using the "summary" function:

```{r}
summary(tablaporfecha)
```

Or it can be calculated using the corresponding function as follows:

```{r}
mean(tablaporfecha$Total_steps)

median(tablaporfecha$Total_steps)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
